% ============================================================
%  ABSTRACT
% ============================================================
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

The convergence of large-scale foundation models with physical robotics represents one
of the most active frontiers in AI research.
This thesis investigates a central question of that frontier: can a zero-shot
vision-language model (VLM), with no robot-specific fine-tuning, serve as the cognitive
core of a physical manipulation agent that reliably completes multi-step tasks in the
real world?

We present the design, implementation, and evaluation of a \textbf{Vision--Language--Action
(VLA) agent} deployed on the PAL Robotics \tiago{} mobile manipulator.
The agent follows a \textbf{ReAct-style perceive-reason-act loop}: at each cycle, it
captures an annotated camera image, assembles a structured scene context, and queries
\textbf{Google Gemini 2.0 Flash} --- a state-of-the-art multimodal LLM --- with a
chain-of-thought-eliciting prompt.
The model responds with a JSON-structured skill selection and natural-language rationale;
the skill is validated against affordance preconditions and executed physically.
Execution outcomes feed back as observations into the next VLM query, closing the loop
in the style of Inner Monologue.

The physical execution layer integrates \textbf{YOLO11n} real-time object detection
(deployed as an HTTP microservice to overcome a Python~3.6 Docker constraint) with a
\textbf{RANSAC cylinder-fitting} 3-D localisation pipeline averaged over five frames
for noise robustness.
The final vertical approach exploits the \tiago{} torso's world-Z-aligned prismatic
joint to guarantee a straight-line end-effector descent without Cartesian IK, a key
insight that emerged through five iterations of real-robot development.
A \textbf{handover skill} closes the ``find--grasp--deliver'' task chain by detecting
a person through YOLO and depth sensing and presenting the object at an ergonomic height.

A \textbf{multi-layer safety system} --- per-skill affordance gating, consecutive
failure tracking, and automatic safe-pose recovery --- prevents unsafe behaviour during
VLM reasoning errors or sensor failures.

Experimental results demonstrate end-to-end task completion of \todo{XX\%} for the full
``find, grasp, and hand over'' scenario, a grasp success rate of \todo{XX\%} over 27
trials, and VLM skill selection accuracy of \todo{XX\%}.
The system confirms that zero-shot frontier VLMs, paired with carefully engineered
classical controllers and affordance feedback, provide a practical and data-free
alternative to large-dataset end-to-end VLA approaches for constrained manipulation
domains.

\vspace{1em}
\noindent\textbf{Keywords:} Embodied AI, Vision-Language-Action Agent, Agentic AI,
Multimodal LLM, Chain-of-Thought, ReAct, Mobile Manipulation, YOLO, Gemini,
Zero-Shot Reasoning, ROS, Grasping, Human-Robot Interaction.
